SDGs

### 2023-10-09 Trilemma_clustering.Rmd
fixed PCA, initial explroations with MFA

```{r fig.width=3, fig.height=3}
pc <- prcomp(df_summary |> 
                 select(-starts_with("country")),
             center = TRUE, scale. = TRUE)
## Fix short names for plotting
df_summary <- df_summary |> 
    left_join(
        dat |> select(country, country_code) |> 
            filter(!is.na(country_code)) |> unique())

pc1 <- ggbiplot::ggbiplot(
    pc, groups = as.character(stab@clusterObjs$kmeans$`3`$cluster),
    ellipse = TRUE, labels = df_summary$country_code, labels.size = 1,
    varname.size = 1.5) +
    scale_color_manual("Groups", values = c("#73B3A3","#FEA621","#5BA4CA")) +
    labs(tag = "A") +
    theme_light(base_size = 5) +
    theme(legend.position = c(0.2, 0.2), panel.grid = element_blank(),
          legend.text = element_text(size = 5), legend.title = element_text(size = 5),
          legend.key.width = unit(3,"mm"), legend.key.height = unit(2, "mm"))

@@

```{r fig.height=3,fig.width=3}
a1 <- df_dat |> 
    as_tibble() |> 
    group_by(country) |> 
    dplyr::summarize(
        EF = sum(d_EF, na.rm = TRUE),
        Gini = sum(d_gini, na.rm = TRUE),
        GNI = sum(d_gni, na.rm = TRUE)
    ) |> 
    mutate(Group = as.character(stab@clusterObjs$kmeans$`3`$cluster)) |> 
    left_join(
        dat |> select(country, country_code) |> 
            filter(!is.na(country_code)) |> unique()) |> 
    ggplot(aes(EF, Gini)) +
    geom_segment(
        aes(x = 0, y = 0, xend = EF, yend = Gini, color = as.factor(Group)),
        arrow = arrow(length = unit(0.1, "cm")), show.legend = FALSE, size = 0.15) +
    geom_text(
        aes(x = EF *1.1, y = Gini * 1.1, label = country_code, colour = Group), 
        size = 1, show.legend = FALSE) +
    scale_color_manual("Groups", values = c("#73B3A3","#FEA621","#5BA4CA")) +
    labs(x = "Ecological Footprint", y = "Inequality (Gini)") +
    theme_light(base_size = 6) +
    theme(panel.grid = element_blank(), legend.position = c(0.8,0.2), 
          legend.text = element_text(size = 4), legend.title = element_text(size =4),
          legend.key.width = unit(2,"mm"), legend.key.height = unit(2, "mm"))

@@

```{r fig.width=3, fig.height=3}
a2 <- df_dat |> 
    as_tibble() |> 
    group_by(country) |> 
    dplyr::summarize(
        EF = sum(d_EF, na.rm = TRUE),
        Gini = sum(d_gini, na.rm = TRUE),
        GNI = sum(d_gni, na.rm = TRUE)
    ) |> 
    mutate(Group = as.character(stab@clusterObjs$kmeans$`3`$cluster)) |> 
    left_join(
        dat |> select(country, country_code) |> 
            filter(!is.na(country_code)) |> unique()) |> 
    ggplot(aes(GNI, Gini)) +
    geom_segment(
        aes(x = 0, y = 0, xend = GNI, yend = Gini, 
            color = as.factor(Group)), show.legend = FALSE,
        arrow = arrow(length = unit(0.1, "cm")), size = 0.15) +
    geom_text(aes(x = GNI * 1.1, y = Gini *1.1, 
                  label = country_code, colour = Group), 
              size = 1, show.legend = FALSE) +
    scale_color_manual("Groups", values = c("#73B3A3","#FEA621","#5BA4CA")) +
    theme_light(base_size = 6) + labs(x = "Prosperity (GNI [log])", y = "Inequality (Gini)") +
    theme(panel.grid = element_blank(), legend.position = c(0.8,0.2),
          legend.text = element_text(size = 4), legend.title = element_text(size = 4),
          legend.key.width = unit(2,"mm"), legend.key.height = unit(2, "mm"))

@@

df_all2 <- df_all |> ungroup() |> 
    filter(!is.na(sd_gini), !is.na(m_ef), !is.na(sd_ef), !is.na(m_gni), !is.na(sd_gni)) 
```

@@

ef_map <-  world |>
    filter(!is.na(pop)) |> 
    left_join(ef, by = c("iso_a2" = "isoa2")) |> 
    ggplot() +
    geom_sf(aes(fill = m_ef), size = 0.05, color = "white") +
    scico::scale_fill_scico(
        "Mean Ecological Footprint",palette = "romaO",
        guide = guide_colorbar(title.position = "top")) + 
    labs(tag = "C") +
    theme_void(base_size = 5) +
    theme(legend.position = "bottom", legend.key.height = unit(1, "mm"), 
          legend.key.width = unit(5, "mm"))

@@

ggsave(
    filename = "pca_maps.png", path = "figures/", device = "png",
    width = 5, height = 4, dpi = 400, bg = "white",
    plot = (pc1 / pc2 ) | (ef_map / gini_map / gni_map) + 
    plot_annotation(tag_levels = "A") + 
    plot_layout(widths = c(1,4), tag_level = "new")
)

>>>>

```{r fig.width=3, fig.height=3}
pc <- prcomp(df_summary |> 
                 select(-starts_with("iso3")),
             center = TRUE, scale. = TRUE)
## Fix short names for plotting
# df_summary <- df_summary |> 
#     left_join(
#         dat |> select(country, country_code) |> 
#             filter(!is.na(country_code)) |> unique())

pc1 <- ggbiplot::ggbiplot(
    pc, groups = as.character(stab@clusterObjs$kmeans$`3`$cluster),
    ellipse = TRUE, labels = df_summary$iso3, labels.size = 1,
    varname.size = 1.5) +
    scale_color_manual("Groups", values = c("#73B3A3","#FEA621","#5BA4CA")) +
    labs(tag = "A") +
    theme_light(base_size = 5) +
    theme(legend.position = c(0.2, 0.2), panel.grid = element_blank(),
          legend.text = element_text(size = 5), legend.title = element_text(size = 5),
          legend.key.width = unit(3,"mm"), legend.key.height = unit(2, "mm"))

@@

```{r fig.height=3,fig.width=3}
a1 <- df_dat |> 
    as_tibble() |> 
    group_by(iso3) |> 
    dplyr::summarize(
        EF = sum(d_EF, na.rm = TRUE),
        Gini = sum(d_gini, na.rm = TRUE),
        GNI = sum(d_gni, na.rm = TRUE)
    ) |> 
    mutate(Group = as.character(stab@clusterObjs$kmeans$`3`$cluster)) |> 
    # left_join(
    #     dat |> select(country, country_code) |> 
    #         filter(!is.na(country_code)) |> unique()) |> 
    ggplot(aes(EF, Gini)) +
    geom_segment(
        aes(x = 0, y = 0, xend = EF, yend = Gini, color = as.factor(Group)),
        arrow = arrow(length = unit(0.1, "cm")), show.legend = FALSE, size = 0.15) +
    geom_text(
        aes(x = EF *1.1, y = Gini * 1.1, label = iso3, colour = Group), 
        size = 1, show.legend = FALSE) +
    scale_color_manual("Groups", values = c("#73B3A3","#FEA621","#5BA4CA")) +
    labs(x = "Ecological Footprint", y = "Inequality (Gini)") +
    theme_light(base_size = 6) +
    theme(panel.grid = element_blank(), legend.position = c(0.8,0.2), 
          legend.text = element_text(size = 4), legend.title = element_text(size =4),
          legend.key.width = unit(2,"mm"), legend.key.height = unit(2, "mm"))

@@

```{r fig.width=3, fig.height=3}
a2 <- df_dat |> 
    as_tibble() |> 
    group_by(iso3) |> 
    dplyr::summarize(
        EF = sum(d_EF, na.rm = TRUE),
        Gini = sum(d_gini, na.rm = TRUE),
        GNI = sum(d_gni, na.rm = TRUE)
    ) |> 
    mutate(Group = as.character(stab@clusterObjs$kmeans$`3`$cluster)) |> 
    # left_join(
    #     dat |> select(country, country_code) |> 
    #         filter(!is.na(country_code)) |> unique()) |> 
    ggplot(aes(GNI, Gini)) +
    geom_segment(
        aes(x = 0, y = 0, xend = GNI, yend = Gini, 
            color = as.factor(Group)), show.legend = FALSE,
        arrow = arrow(length = unit(0.1, "cm")), size = 0.15) +
    geom_text(aes(x = GNI * 1.1, y = Gini *1.1, 
                  label = iso3, colour = Group), 
              size = 1, show.legend = FALSE) +
    scale_color_manual("Groups", values = c("#73B3A3","#FEA621","#5BA4CA")) +
    theme_light(base_size = 6) + labs(x = "Prosperity (GNI [log])", y = "Inequality (Gini)") +
    theme(panel.grid = element_blank(), legend.position = c(0.8,0.2),
          legend.text = element_text(size = 4), legend.title = element_text(size = 4),
          legend.key.width = unit(2,"mm"), legend.key.height = unit(2, "mm"))

@@

df_all2 <- df_all2 |> ungroup() |> 
    filter(!is.na(sd_gini), !is.na(m_ef), !is.na(sd_ef), !is.na(m_gni), !is.na(sd_gni))

save(df_all2, file = "data/trilemma_static.Rda")
```

@@

ef_map <-  world |>
    filter(!is.na(pop)) |> 
    left_join(ef, by = c("iso_a2" = "isoa2")) |> 
    ggplot() +
    geom_sf(aes(fill = m_ef), size = 0.05, color = "white") +
    scico::scale_fill_scico(
        "Mean Ecological Footprint",palette = "romaO", direction = -1,
        guide = guide_colorbar(title.position = "top")) + 
    labs(tag = "C") +
    theme_void(base_size = 5) +
    theme(legend.position = "bottom", legend.key.height = unit(1, "mm"), 
          legend.key.width = unit(5, "mm"))

@@

ggsave(
    filename = "pca_maps.png", path = "figures/", device = "png",
    width = 5, height = 4, dpi = 600, bg = "white",
    plot = (pc1 / pc2 ) | (ef_map / gini_map / gni_map) + 
    plot_annotation(tag_levels = "A") + 
    plot_layout(widths = c(1,4), tag_level = "new")
)

### 2023-10-09 inequality-SDGs.Rmd
fixed PCA, initial explroations with MFA

```{r}
ineq_dat
```
`m` is total, `a` is average. All other codes are documented here: [https://wid.world/codes-dictionary/#packages](https://wid.world/codes-dictionary/#packages).

@@

```{r}
## both datasets are wider format, so makes sense to merge as is
df_dat <- df_dat |>
    left_join(countries |> select(country_code, country = x2_alpha_code)) |> 
    left_join(ineq_dat |> 
                  select(-age, -pop) |> 
                  unite(col = "var", c("variable", "percentile")) |> 
                  pivot_wider(names_from = var, values_from = value)
              )
```

@@

df_un_reduced <- df_un_reduced |> ungroup() |> #group_by(iso3)
    left_join(
        countries |> select(iso3 = country_code, country = x2_alpha_code) ) |> 
    # filter(is.na(country)) |> pull(iso3) |> unique()
    unite(col = "var", year, series_name, sep = "_") |> 
    pivot_wider(values_from = value, names_from = var) |> 
    left_join(
        ineq_dat |> 
            select(-age, -pop) |> 
            filter(year >=  2000, year < 2019) |> # max year 2018 in UN dataset
            unite(col = "series_name", c("variable", "percentile")) |> 
            right_join(
                countries |> 
                    select(iso3 = country_code, country = x2_alpha_code) |> 
                    filter(iso3 != "CHI") # filter out channel islands
            ) |> 
            unite(col = "var", year, series_name, sep = "_") |> 
            pivot_wider(values_from = value, names_from = var) 
        )
```

@@

```{r}
#save(df_dat, df_un_reduced, file = "data/cleaned_sdg.RData")
```

@@

```{r}
# some notes on the pca object
pca1$rotation |> dim() # 171 variables*years, 160 PCs because there was 160 countries to start with.
pca1$x |> dim() # The loadings of 160 countries in each of the 160 components.
pca1$center |> length() # 171 variable centers
pca1$scale |> length() # 171 variable scales.
```

@@

pca2 <- df_un_reduced |>
    ungroup() |> 
    select(-country, -iso3, -NA_NA) |> #skimr::skim()
    select(-contains("sptinc"), -contains("gptinc"), -contains("rptinc")) |> 
    prcomp(center = TRUE, scale. = TRUE)

```
**J230529:** It doesn't work, there is no enough variability in the data. It seems I have several variables / countries with zero variance or constants despite no missing values. However, it only screws up when adding the inequality variables. Without it does still run, why? Check with Maike. It might have to do with number of observations (countries) vs number of variables. Most likely too wide of a dataset can be problematic.

```{r}
biplot(pca2)
```
### Multiple factor analysis

```{r}
mfa1 <- df_un_reduced |> 
    select(-iso3, -country, -NA_NA) |> 
    as.matrix(dimnames = list(df_un_reduced |> pull(iso3) |> unique())) |> 
    FactoMineR::MFA(group = rep(19, 10), type = rep("s", 10)) # 18 years times 6 variables
```

@@

```{r}
mfa1 |> summary()
```

@@

@@

```{r}
pca3 <- df_dat |> 
    ungroup() |> 
    mutate(year = as_factor(year)) |> 
    split(~year) |>
    map(function(x) select(x, where(is.numeric))) |> 
    map(function(x) princomp(x, center = TRUE, scale. = TRUE))
```

```{r}
pca3 |> map(biplot)
```

>>>>

```{r}
ineq_dat

## correct Namibia is not a NA, but the country code is NA
ineq_dat <- ineq_dat |> 
    mutate(country = case_when(is.na(country) ~ "NA", TRUE ~ country)) #|> filter(country == "NA")
```
`m` is total, `a` is average. All other codes are documented here: [https://wid.world/codes-dictionary/#packages](https://wid.world/codes-dictionary/#packages).

@@

```{r}
## both datasets are wider format, so makes sense to merge as is
df_dat <- df_dat |>
    left_join(countries |> select(country_code, country = x2_alpha_code)) |> 
    left_join(ineq_dat |> 
                  select(-age, -pop) |> 
                  unite(col = "var", c("variable", "percentile")) |> 
                  pivot_wider(names_from = var, values_from = value)
              )

df_dat
```

@@

df_un_reduced <- df_un_reduced |> ungroup() |> #group_by(iso3)
    left_join( 
        countries |> select(iso3 = country_code, country = x2_alpha_code) ) |> 
    filter(iso3 != "FJI") |> # drop out Fiji, it does not exist in ineq data and creates NAs
    #filter(is.na(country)) |> pull(iso3) |> unique()
    unite(col = "var", year, series_name, sep = "_") |> 
    pivot_wider(values_from = value, names_from = var) |> 
    left_join(
        ineq_dat |> 
            select(-age, -pop) |> 
            filter(year >=  2000, year < 2019) |> # max year 2018 in UN dataset
            unite(col = "series_name", c("variable", "percentile")) |>
            filter(series_name != "sptinc992j_p0p100") |> # all ones, no variability
            right_join(
                countries |> 
                    select(iso3 = country_code, country = x2_alpha_code) |> 
                    filter(iso3 != "CHI") # filter out channel islands
            ) |> 
            unite(col = "var", year, series_name, sep = "_") |> 
            pivot_wider(values_from = value, names_from = var) |> 
            select(-NA_NA)
        )

### No missing value in any variable!
df_un_reduced |>
    pivot_longer(3:last_col(), names_to = "variable", values_to = "value") |> 
    mutate(
        year = str_sub(variable, 1L, 4L),
        variable = str_sub(variable, 6L, -1L)
    ) |> 
    filter(is.na(value))

@@

```{r}
# save(df_dat, df_un_reduced, file = "data/cleaned_sdg.RData")
```

@@

```{r}
# some notes on the pca object
pca1$rotation |> dim() # 228 variables*years, 150 PCs because there was 150 countries after filtering missing values in rptinc992j_p0p100.
pca1$x |> dim() # The loadings of 150 countries in each of the 150 components.
pca1$center |> length() # 228 variable centers
pca1$scale |> length() # 228 variable scales.
```

@@

pca2 <- df_un_reduced |>
    select(-country, -iso3) |>
    prcomp(center = TRUE, scale. = TRUE)
```
**J230529:** It doesn't work, there is no enough variability in the data. It seems I have several variables / countries with zero variance or constants despite no missing values. However, it only screws up when adding the inequality variables. Without it does still run, why? Check with Maike. It might have to do with number of observations (countries) vs number of variables. Most likely too wide of a dataset can be problematic.

**J231009** Now it's working, the problem was not the zero or low variance, the problem was NA induced by two countries that didn't match on the joints, Namibia whose iso2 is NA and got interpreted as missing value, and Fiji that does not exist on the WII database. So only 50 countries at the end.

```{r}
biplot(pca2)
```

```{r}
x <- summary(pca2, loadings = F, cutoff = 0.1)
x <- x$importance |> as_tibble(rownames = "var") |> 
    pivot_longer(cols = starts_with("PC"), names_to = "comp") |> 
    mutate(comp = str_remove(comp, "PC") |> as.numeric()) |> 
    dplyr::filter( comp <= 10 , var != "Standard deviation")

a <- pca2$x |> as_tibble() |> 
    select(PC1:PC2) |> 
    mutate(iso = df_un_reduced |> pull(iso3) |> unique() ) |> 
    ggplot(aes(PC1, PC2)) +
    geom_text(aes(label = iso), size = 2) +
    coord_equal() +
    theme_light(base_size = 6)

b <- ggplot() +
    geom_segment(
        data = pca2$rotation |>
            as_tibble(rownames = "factor") |> 
            mutate(year = str_sub(factor, start = 1L, end = 4L),
                   factor = str_remove(factor, pattern = "\\d{4}_"))  |> 
            select(year, factor, PC1:PC2),
        aes(xend = PC1, yend = PC2, color = factor, alpha = as.numeric(year)),
        x = 0, y = 0,
        arrow = arrow(length = unit(0.1,"cm")), linewidth = 0.3) +
    scale_color_brewer("Variable",palette = "Set3",
        guide = guide_legend(ncol = 2, title.position = "top")) +
    scale_alpha("Year", guide = guide_legend(ncol = 1, title.position = "top")) +
    xlim(c(-0.1, 0.15)) +
    theme_light(base_size = 6)

### Multiple factor analysis

```{r}
df_un_reduced |> 
    pivot_longer(3:last_col(), names_to = "variable", values_to = "value") |> 
    separate(variable, into = c("year", "indicator"), sep = "_") |> 
    skimr::skim() # 19 years, 9 variables

mfa1 <- df_un_reduced |> 
    select(-iso3, -country) |> 
    as.matrix(dimnames = list(df_un_reduced |> pull(iso3) |> unique())) |> 
    FactoMineR::MFA(group = rep(19, 9), type = rep("s", 9)) # 19 years times 9 variables
```

@@

```{r}
# df_dat$year |> unique() |> length() # 19yrs

mfa2 <- df_dat |> 
    select(-country_name, -country, -sptinc992j_p0p100) |> 
    filter(!is.na(rptinc992j_p0p100)) |> 
    pivot_longer(cols = `Access to electricity (% of population)`:last_col(),
                 values_to = "value", names_to = "variable") |> #pull(variable) |> unique() 12 vars
    unite(col = "var", year, variable, sep = "_") |> 
    pivot_wider(names_from = var, values_from = value) |> 
    ungroup() |> 
    select(-country_code) |> 
    as.matrix(dimnames = list(df_dat |> pull(country_code) |> unique())) |> 
    FactoMineR::MFA(group = rep(19, 12), type = rep("s", 12)) # 19 years times 9 variables
```

@@

```{r}
ggbiplot::ggbiplot(
    pca2,
    labels = df_un_reduced |> pull(iso3) |> unique(),
    circle = TRUE,
    varname.abbrev = TRUE)
```

Extract the weight per country on the relevant components to see how they move on PCA or MFA space.

```{r}
mfa1 |> summary()
mfa2$summary.quanti
```

@@

```{r}
# pca3 <- df_dat |> 
#     ungroup() |> 
#     mutate(year = as_factor(year)) |> 
#     split(~year) |>
#     map(function(x) select(x, where(is.numeric))) |> 
#     map(function(x) princomp(x, center = TRUE, scale. = TRUE))
```

```{r}
# pca3 |> map(biplot)
```
